---
title: "Cryptocurrency Volatility Forecasting"
author: "Elliot Yang"
date: today
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---

## Executive summary

This project develops an automated cryptocurrency volatility forecasting framework using GARCH-family models for Bitcoin (BTC), Ethereum (ETH), Solana (SOL), and Binance Coin (BNB). Daily log returns are analysed using three GARCH-family specifications: GARCH(1,1), GJR-GARCH(1,1), and EGARCH(1,1). The best model for each cryptocurrency is selected based on AIC, BIC, and log-likelihood before producing out-of-sample volatility forecasts. Forecast performance is evaluated using RMSE, MAE, and Value-at-Risk (VaR), with all analyses generated automatically through a reproducible Quarto workflow.

## Introduction

This project develops an automated cryptocurrency volatility forecasting framework using GARCH-family models.

The workflow consists of:

```         
Yahoo Finance
      │
      ▼
Download Prices
      │
      ▼
Calculate Log Returns
      │
      ▼
ADF Test
      │
      ▼
ARCH Test
      │
      ▼
Estimate GARCH Models
      │
      ▼
Model Selection
      │
      ▼
Forecast Volatility
      │
      ▼
RMSE • MAE • VaR
```

## Data overview

### Cryptocurrency Returns

```{r}
library(tidyverse)
returns <- read_csv("../data/processed/cleaned_returns.csv")

glimpse(returns)
```

```{r}
head(returns)
```

## Best model

### Selected Models

```{r}
best_models <- readRDS("../outputs/models/best_models.rds")

data.frame(
  Cryptocurrency = names(best_models),
  Model = sapply(best_models,
                 function(x) x@model$modeldesc$vmodel)
)
```

## Forcast accuracy

```{r}
accuracy <- read_csv("../outputs/tables/forecast_accuracy.csv")

knitr::kable(accuracy)
```

## VaR

```{r}
var <- read_csv("../outputs/tables/var_results.csv")

knitr::kable(var)
```

## Forecast plots

```{r}
library(ggplot2)
library(dplyr)
library(readr)
library(rugarch)
returns <- read_csv(
  "../data/processed/cleaned_returns.csv",
  show_col_types = FALSE
)

best_models <- readRDS("../outputs/models/best_models.rds")

coins <- c("BTC", "ETH", "SOL", "BNB")

split <- floor(0.8 * nrow(returns))

for (coin in coins) {

  fit <- best_models[[coin]]


  forecast_df <- read_csv(
    paste0("../outputs/forecasts/", coin, "_forecast.csv"),
    show_col_types = FALSE
  )
  hist_vol <- tibble(
  Date = returns$Date,
  Volatility = as.numeric(sigma(fit)),
  Type = "Historical"
)
  forecast_vol <- tibble(
  Date = seq(
    from = max(hist_vol$Date) + 1,
    by = "day",
    length.out = nrow(forecast_df)
  ),
  Volatility = forecast_df$Forecast_Volatility,
  Type = "Forecast"
)

  plot_df <- bind_rows(
    tail(hist_vol, 300),
    forecast_vol
)
  
  forecast_start <- max(hist_vol$Date)

  p <- ggplot(plot_df,
              aes(x = Date,
                  y = Volatility,
                  colour = Type)) +

    geom_line(linewidth = 1) +

    geom_vline(
      xintercept = forecast_start,
      linetype = "dashed",
      colour = "grey50"
    ) +

    scale_colour_manual(
      values = c(
        Historical = "#1F4E79",
        Forecast = "#D62728"
      )
    ) +

    labs(
      title = paste(coin, "Conditional Volatility"),
      subtitle = "Historical fitted volatility and forecasted conditional volatility",
      x = "Date",
      y = "Conditional Volatility",
      colour = ""
    ) +

    theme_minimal(base_size = 13) +

    theme(
      legend.position = "top",
      plot.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )

  print(p)
  ggsave(
  filename = paste0("../figures/", coin, "_volatility_forecast.png"),
  plot = p,
  width = 8,
  height = 5,
  dpi = 300
)
}



```

## Conclusion

This project successfully demonstrates an automated cryptocurrency volatility forecasting framework using GARCH-family models. Statistical testing confirmed the presence of volatility clustering, supporting the use of conditional variance models. After comparing GARCH(1,1), GJR-GARCH(1,1), and EGARCH(1,1), the best-performing specification was selected for each cryptocurrency. The resulting forecasts, together with RMSE, MAE, and VaR measures, provide a reproducible framework for analysing and managing cryptocurrency market risk.
